---
title: "Time Series Analysis - Singapore Total Fertility Rate\ 
  and Total Live Births"
author: "Hunter Holt"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: false
    toc_depth: 3
    number_sections: false
    df_print: kable
    highlight: tango
    latex_engine: pdflatex
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{ragged2e}
  - \usepackage{parskip}
  - \usepackage{setspace}
  - \usepackage{titlesec}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \setlength{\parindent}{0pt}
  - \setlength{\parskip}{6pt}
  - \titleformat{\section}{\normalfont\bfseries}{\thesection}{1em}{}
  - \titleformat{\subsection}{\normalfont\bfseries}{\thesubsection}{1em}{}
  - \titlespacing*{\section}{0pt}{12pt}{4pt}
  - \titlespacing*{\subsection}{0pt}{8pt}{2pt}
  - \renewcommand{\arraystretch}{1.15}
bibliography: references.bib
---

```{r setup, include=FALSE, warnings=FALSE}
knitr::opts_chunk$set(
	echo		= FALSE,
	warning	= FALSE,
	message		= FALSE,
	fig.align	= "center",
	fig.pos = "H",
  out.width = "85%"
)
```


```{r libs, include=FALSE}
library(tidyverse)	# for data wrangling & ggplot2
library(tseries)	# for adf & kpss tests
library(forecast)	# for acf/pacf, forecast & checkresiduals
library(tsibble)	# for tsibble data structure
library(knitr)
library(kableExtra)
```

---

\begin{center}
\large\textbf{Abstract}
\end{center}

Over the course of the last century, Singapore's demographic transition remains as one of the most rapid ever recorded, with the Total Fertility Rate (TFR) declining from 5.76 in 1960 to less than 1.0 by 2024. This paper features a complete time series analysis of two interrelated population metrics: TFR and Total Live Births (TLB). These are scraped from the Singapore Department of Statistics' annual data, ranging from 1968 to 2024. After an established Data Analysis Lifecycle, paired with the Box-Jenkins identification cycle, a log-differencing transformation is required to achieve stationarity. This is also furthermore confirmed by ADF and KPSS tests. Autocorrelation (ACF) and partial autocorrelation (ACF) functions applied to our stationary series prompts for an ARIMA$(p, 1, q)$ candidate model(s) fitted with a logarithmic Box-Cox transform. Model selection verification using AICc identifies the specification with best fit for each series; residual diagnostics aim to replicate white noise behaviour. Out-of-sample forecasts over a subsample of 2013--2024 are evaluated against observed values, with performance assessed via RMSE, MAE, MAPE, and Theil's U. The results post-evaluation confirm the particular difficulty of accurately forecasting sub-replacement fertility prediction mapping, likely due to structural policy interventions within Singapore and/or the impact of COVID-19. This motivates additional extension processes including ARIMA-GARCH and state-space modelling.

*Keywords:* Total Fertility Rate; Total Live Births; ARIMA; Box-Jenkins; Stationarity; Singapore

\vspace{10pt}
\hrule
\vspace{8pt}

---

# 1 Introduction

- Demographic transition
- Discuss variables
- Discuss significant change in variables, and why. EDA
- Introduce research questions: 'which policy interventions over 1960--2024 justify the significant structural deviations in TFR and TLB?', 'can ARIMA models trained on the 1960--2012 window project the 2013--2024 period?', 'what (and how do) demographic mechanisms account for the recent significant drop in TFR?'
- State IED Box-Jenkins cycle as diagnostic tool

# 2 Data

# 2.1 Data Sourcing and Structuring

```{r load-data}
raw <- read.csv("SingaporeCensus.csv", 
                check.names = FALSE, 
                na.strings = c("", 
                               "NA", 
                               "na")
)

# Transpose
long <- raw |>
    rename(Series = DataSeries) |>
    pivot_longer(cols = -Series,
                 names_to = "Year",
                 values_to = "Value", 
                 names_transform = list(Year = as.integer)
                 ) |>
    filter(Year >= 1960, Year <= 2024)
	
# Extract TFR and TLB data
tfr_df <- long |>
	filter(trimws(Series) == "Total Fertility Rate (TFR)") |>
	arrange(Year)
tlb_df <- long |>
	filter(trimws(Series) == "Total Live-Births") |>
	arrange(Year)
	
# Verify 65 years covered per dataset
stopifnot(nrow(tfr_df) == 65, nrow(tlb_df) == 65)

# Create cleaned tsibbles
tfr_tsb <- ts(tfr_df$Value, start = 1960, frequency = 1)
tlb_tsb <- ts(tlb_df$Value, start = 1960, frequency = 1)

# Split tsibble for training & testing
tfr_train <- window(tfr_tsb, end = 2012)
tfr_test <- window(tfr_tsb, start = 2013)
tlb_train <- window(tlb_tsb, end = 2012)
tlb_train <- window(tlb_tsb, start = 2013)
```

The data used throughout this report are sourced from *SingaporeCensus.csv*, a population census published by the Singapore Department of Statistics.
The raw comma-separated-value file is stored in wide format, i.e. the years 1960--2024 are column headers. Two rows are extracted -- one for TFR and one for TLB -- and reshaped to long format by using `pivot_longer()`. The resulting cleaned dataset contains 65 annual observations per series with no missing values.

For forecasting evaluation, the full series is partitioned into a training windows of $n = 53$ observations (from 1960 to 2012) and a complementary windows of $h = 12$ observations (2013 to 2024).

# 2.2 Statistical Summary

```{r summary-table}
summary_tbl <- tibble(
                      Statistic = c("Min", "Q1", "Median", "Mean", "Q3", "Max", "NAs"),
                      TFR = c(min(tfr_tsb), quantile(tfr_tsb, .25), median(tfr_tsb), mean(tfr_tsb), quantile(tfr_tsb, .75), max(tfr_tsb), sum(is.na(tfr_tsb))),
                      TLB = c(min(tlb_tsb), quantile(tlb_tsb, .25), median(tlb_tsb), mean(tlb_tsb), quantile(tlb_tsb, .75), max(tlb_tsb), sum(is.na(tlb_tsb)))
                      )
kable(summary_tbl, digits = 2, booktabs = TRUE, caption = "Summary statistics for TFR and TLB (1960--2024).") |>
  kable_styling(latex_options = c("hold_position", "striped"), font_size = 10)
```

TFR ranges from an initial post-war high of 5.76, to a final minimum of 0.97, which is much lower than the 'replacement threshold' of 2.1 [@Espenshade2003]. The considerable difference between the mean of 2.06 and median of 1.62 is reflective of a strongly right-skewed distribution produced by the pre-transitional years.

# 2.3 Time Series Plots

```{r ts-plots, fig.width=8.5, fig.height=3.8, fig.cap="TFR (left) and TLB (right), from 1960--2024. The dashed vertical line indicates the train-test data partitioning."}
par(mfrow = c(1, 2), mar = c(4, 4.2, 2.5, 1), family = "serif")
plot(tfr_tsb, 
     main = "Total Fertility Rate", 
     ylab = "TFR", xlab = "Year",
     col = "black", lwd = 1.2, las = 1
     )
abline(v = 2012, 
       lty = 2, 
       col = "grey55"
       )
plot(tlb_tsb,
     main = "Total Live Births",
     ylab = "TLB", xlab = "Year",
     col = "black", lwd = 1.2, las = 1
     )
abline(v = 2012,
       lty = 2,
       col = "grey55"
       )
par(mfrow = c(1, 1))
```

Figure 1 indicates two salient features shared by both series. For one, a strong uniform downward trend persists over the full sample, which immediately rules out non-transformative modelling approaches. For two, the variance of the fluctuations around the trend is considerably higher during earlier years (1960--1980) instead of during the sub-replacement period post-1985. This is indicative of a multiplicative error structure rather than an additive one, which prompts for a logarithmic variance stabilisation method.

# 3 Context and Methodology

## 3.1 Transformation and Stationarity

A time series $\{X_t\}$ is said to be *weakly stationary* if its unconditional mean $\mu = E[X_t]$ is constant and its autocovariance function $\gamma(h) = \mathrm{Cov}(X_t, X_{t-h})$ depends only on lag $h$, rather than time $t$ [@BoxJenkins1970]. Both raw series are in violation of this condition; shown is a persistent downward trend (i.e. a non-constant mean) and heteroskedastic fluctuations (i.e. non-constant variance).

Since variance appears proportional to level, a logarithmic transformation $W_t = \log X_t$ is applied to stabilise variance. The log series retains a trend, prompting for a single first-difference to produce a stationary working series:

$$
Y_t = (1 - B)\log X_t = \log X_t - \log X_{t-1} = \log\!\left(\frac{X_t}{X_{t-1}}\right),
$$

where $B$ is the backward shift operator ($BX_t = X_{t-1}$). In this context, $Y_t$ is interpreted as the log-annual growth rate of the series. Stationarity of $Y_t$ is assessed using two statistical tests:

- **Augmented Dickey-Fuller (ADF)** [@Dickney1979]: tests $H_0$ (unit root) against $H_1$ (stationary). A small $p$-value supports rejection of the unit-root null.
- **KPSS** [@Kwiatkowski1992]: tests $H_0$: (trend-)stationary against $H_1$ (unit root). A large $p$-value (non-rejection) supports the stationarity hypothesis. 

Using both tests simultaneously ensures less ambiguity that individual tests can experience; due to low power or size distortion, for instance.

## 3.2 ARIMA Model specification

An ARIMA$(p, d, q)$ model applied to the log-transformed series $\log X_t$ (so that $d = 1$ acts on the log series) is:

$$
\phi(B)(1 - B)\log X_t = c + \theta(B)\varepsilon_t, \qquad \varepsilon_t \sim \mathrm{WN}(0, \sigma^2),
$$

where $\mathrm{WN}(0, \sigma^2)$ denotes a white-noise sequence (an uncorrelated process with zero mean and constant variance $\sigma^2$). The AR and MA polynomials are:

$$
\phi(B) = 1 - \phi_1 B - \phi_2 B - \phi_2 B^2 - \cdots \phi_p B^p, \qquad \theta(B) = 1 + \theta_1 B + \theta_2 B^2 + \cdots + \theta_q B^q.
$$

This model is *causal* (stationary) if all roots of $\phi(B) = 0$ lie strictly outside the unit circle; additionally, it is *invertible* if all roots of $\theta(B) = 0$ lie strictly outside the unit circle.

To specify this, the `forecast` R package features `Arima(x, order = c(p, 1, q), lambda = 0)`, where `lambda = 0` applies the logarithmic Box-Cox transform internally, and automatically back-transforms forecasts to the original scale.

### 3.2.1 Model identification

Given a stationary series $Y_t$, the **autocorrelation function** (ACF) at lag $k$ is:

$$
\rho_k = \frac{\mathrm{Cov}(Y_t,\,Y_{t-k})}{\mathrm{Var}(Y_t)},
$$

and the **partial autocorrelation function** (PACF) $\alpha_k$ measures the linear dependence between $Y_t$ and $Y_{t-k}$ after eliminating the effect pf $Y_{t-1}, \ldots, Y_{t-k+1}$. The Box-Jenkins identification table is:

\begin{table}
Model & Theoretical ACF & Theoretical PACF \\
\hline
AR$(p)$ & Geometric decay/oscillation & Cuts off after lag $p$ \\
MA$(q)$ & Decays after lag $q$ & Geometric decay/oscillation \\
ARMA$(p, q)$ & Decays after lag $q - p$ & Decays after lag $p - q$
\end{table}

Approximate 95\% bounds on the sample ACF/PACF are $\pm 1.96/\sqrt{n}$, where $n$ is the length of the differenced series.

### 3.2.2 Model selection

Candidate models are compared using the corrected Akaike Information Criterion:

$$
\mathrm{AICc} = -2\,\hat{\ell} + 2k + \frac{2k(k + 1)}{n - k - 1},
$$

where $\hat{\ell}$ is the maximised log-likelihood and $k$ is the number of free parameters. AICc is typically preferred over AIC for small $n$ [@BurnhamAnderson2002] as a stronger model complexity penalty reduces the risk of overfit and over-parameterisation.

## 3.3 Residual Diagnostics

An accurate, well-specified model should yield residuals $\hat{\varepsilon}_t$ that resemble white noise. This is assessed using three complementary tools:

1. **ACF of residuals**. All sample autocorrelations should lie within the 95% (i.e., $\pm 1.96/\sqrt{n}$);
2. **Ljung-Box test** [@Ljung-Box1978]. The portmanteau statistic $$Q^* = n(n + 2)\sum_{k=1}^m \frac{\hat{\rho}_k^2}{n - k}$$ is compared to $\chi^2(m - p - q)$ under $H_0$ (i.e., no residual corelation prior to lag $m$). The degrees of freedom correction `fitdf` $= p + q$ accounts for parameters that are influenced by estimation error. Omitting it (setting `fitdf = 0`) artifically 'inflates' the $\chi^2$ test statistic;
3. **Residual histogram**: valid Gaussian prediction intervals require residuals that are approximately Normal.

## 3.4 Forecast Accuracy Measurement

Point-forecast performance over our 'hold-out' period (2013--2024) is measured by:

$$
\mathrm{RMSE} = \sqrt{\frac{1}{h}\sum_{t = 1}^h e_t^2}, \quad \mathrm{MAE} = \frac{1}{h}\sum_{t=1}^h |e_t|, \quad \mathrm{MAPE} = \frac{100}{h}\left|\frac{e_t}{X_t}\right|
$$

where $e_t \hat{X}_t - X_t$. As mentioned, **Theil's U** compares model RMSE to that of the naive (that is, the random walk) forecast; $U < 1$ indicates the model outperforms the initial naive benchmark.

---

# 4 Exploratory Analysis

## 4.1 Log Transformation Assessment

```{r log-plots, fig.width=8.5, fig.height=4.5, fig.cap="Raw versus log-transformed series for TFR (top row) and TLB (bottom row); training period 1960--2012."}
par(mfrow = c(2, 2), mar = c(4, 4, 2.5, 1), family = "serif")
plot(tfr_train, main = "TFR - Raw", ylab = "TFR", xlab = "", las = 1)
plot(log(tfr_train), main = "TFR - Log", ylab = "log(TFR)", xlab = "", las = 1)
plot(tlb_train, main = "TLB - Raw", ylab = "Live Births", xlab = "Year", las = 1)
plot(log(tlb_train), main = "TLB - Log", ylab = "log(TLB)", xlab = "Year", las = 1)

par(mfrow = c(1, 1))
```

Figure 2 confirms that the log-transformation reduces heteroskedasticity considerably among both series; the large fluctuations earlier on in TLB are quite attenuated in comparison to the stable post-1985 period. However, a downward trend continues among both series, which prompts for first differencing.

# 4.2 Log-Differenced Series

```{r logdiff-series, fig.width=8.5, fig.height=3.5, fig.cap="Log-differenced series $(1-B)\\log X_t$ for TFR (left) and TLB (right), over training period 1961--2012."}
d_tfr <- diff(log(tfr_train))
d_tlb <- diff(log(tlb_train))

par(mfrow = c(1, 2), mar = c(4, 4.2, 2.5, 1), family = "serif")
plot(d_tfr, main = expression((1-B)*log(TFR)), ylab = "Log-difference", xlab = "Year", las = 1)
abline(h = 0, lty = 2, col = "grey55")
plot(d_tlb, main = expression((1-B)*log(TLB)), ylab = "Log-difference", xlab = "Year", las = 1)
abline(h = 0, lty = 2, col = "grey55")
par(mfrow = c(1, 1))
```

## 4.3 Stationary Tests

### 4.3.1 Raw Series

```{r stationarity-raw}
adf_tfr <- adf.test(tfr_train)
kpss_tfr <- kpss.test(tfr_train, null = "Level")

adf_tlb <- adf.test(tlb_train)
kpss_tlb <- kpss.test(tlb_train, null = "Level")

stationarity_raw <- tibble(
	Series = c("TFR", "TLB"),
	`ADF p-value` = c(round(adf_tfr$p.value, 4), round(adf_tlb$p.value, 4)),
	`KPSS p-value` = c(round(kpss_tfr$p.value, 4), round(kpss_tlb$p.value, 4)),
  Conclusion = c("Non-stationary", "Non-stationary")
)
kable(stationarity_raw, booktabs = TRUE, caption = "Stationarity tests on raw training series (1960--2012).") |>
  kable_styling(latex_options = c("hold_position", "striped"), font_size = 10)
```

### 4.3.2 Log-Differenced series

```{r stationarity-logdiff}
adf_d_tfr <- adf.test(d_tfr)
kpss_d_tfr <- kpss.test(d_tfr, null = "Level")

adf_d_tlb <- adf.test(d_tlb)
kpss_d_tlb <- kpss.test(d_tlb, null = "Level")

stationarity_diff <- tibble(
	Series = c("$(1-B)\\log(\\text{TFR})$", "$(1-B)\\log(\\text{TLB})$"),
  `ADF p-value` = c(round(adf_d_tfr$p.value, 4), round(kpss_d_tlb$p.value, 4)),
  Conclusion = c("Stationary", "Stationary")
  )
kable(stationarity-diff, booktabs = TRUE, escape = FALSE, caption = "Stationarity tests on log-differenced training series.") |>
  kable_styling(latex_options = c("hold_position", "striped"), font_size = 10)
```


## 4.4 ACF & PACF Analysis

```{r acf-pacf-logdiff, fig.width = 8.5, fig.height = 4.5, fig.cap="ACF (left column) and PACF (right column) of $(1-B)\\log X_t$ for TFR (top row) and TLB (bottom row), over training period. Lags extend to 25. Blue dasheed lines are 95\\% bounds $\\pm 1.96/\\sqrt{n}$."}
par(mfrow = c(2, 2), mar = c(4, 4.2, 3, 1), family = "serif")

Acf(d_tfr, lag.max = 25, main = expression(Delta * log(TFR) ~ "-- ACF"))
Pacf(d_tfr, lag.max = 25, main = expression(Delta * log(TFR) ~ "-- PACF"))
Acf(d_tlb, lag.max = 25, main = expression(Delta * log(TLB) ~ "-- ACF"))
Pacf(d_tlb, lag.max = 25, main = expression(Delta * log(TLB) ~ "--PACF"))

par(mfrow = c(1, 1))
```

# 5. Model Identification and Estimation

## 5.1 TFR Candidate Model

- **T1**: ARIMA$(0, 1, 1)$; pure MA(1) on the log-differenced series;
- **T2**: ARIMA$(1, 1, 1)$; ARMA(1, 1) on the log-differenced series;
- **T3**: ARIMA$(1, 1, 0)$; pure AR(1) on the log-differenced series.

```{r fit-tfr, result='hide'}
fit_T1 <- Arima(tfr_train, order = c(0, 1, 1), lambda = 0, include.constant = TRUE)
fit_T2 <- Arima(tfr_train, order = c(1, 1, 1), lambda = 0, include.constant = TRUE)
fit_T3 <- Arima(tfr_train, order = c(1, 1, 0), lambda = 0, include.constant = TRUE)
```

```{r tfr-model-selection}
tfr_sel <- tibble(
                  Model = c("ARIMA(0, 1, 1)", "ARIMA(1, 1, 1)", "ARIMA(1, 1, 0)"),
                  AIC = c(AIC(fit_T1), AIC(fit_T2), AIC(fit_T3)),
                  AICc = c(fit_T1$aicc, fit_T2$aicc, fit_T3$aicc),
                  BIC = c(BIC(fit_T1), BIC(fit_T2), BIC(fit_T3))
                  )

best_T <- which.min(tfr_sel$AICc)
kable(tfr_sel, digits = 3, booktabs = TRUE, caption = "Information criteria for TFR candidate models (log scale).") |>
  kable_styling(latex_options = c("hold_position", "striped"), font_size = 10) |>
  row_spec(best_T, bold = TRUE)
```

```{r tfr-select-fit}
tfr_names <- c("ARIMA(0, 1, 1)", "ARIMA(1, 1, 1)", "ARIMA(1, 1, 0)")
tfr_fits <- list(fit_T1, fit_T2, fit_T3)
tfr_best <- tfr_fits[[best_T]]
tfr_best_name <- tfr_names[best_T]
```

The model with the lowest AICc (bolded in Table 4, for reference) is selected. Its estimated coefficients are printed below:
```{r tfr-coef-table}
cf <- coef(tfr_best);   se <- sqrt(diag(vcov(tfr_best)))
zv <- cf / se;          pv <- 2 * pnorm(-abs(zv))
kable(
      tibble(Parameter = names(cf),
             Estimate = round(cf, 5),
             `Std. Error` = round(se, 5),
             `z-stat` = round(zv, 3),
             `p-value` = round(pv, 4)),
      booktabs = TRUE,
      caption = paste("Estimated parameters:", tfr_best_name, "(log-scale TFR).")
      ) |>
kable_styling(latex_options = c("hold_position", "striped"), font_size = 10)
```

\newpage
## TLB - Candidate Model

```{r model-tlb, echo=FALSE, fig.width=6, fig.height=5, results='hide'}
tlb_model1 <- Arima(tlb_train_ts, order = c(0, 0, 0)) # TBD
```

```{r model-tlb-results, echo=FALSE}
lb_tlb <- Box.test(residuals(tlb_model1), lag = 10, type = "Ljung-Box", fitdf = 0)

knitr::kable(
	accuracy(tlb_model1) |> as.data.frame(),
	digits = 4, caption = "TLB training set error measures"
)

knitr::kable(
	tibble(
		'Q*' = lb_tlb$statistic,
		df = lb_tlb$parameter,
		'p-value' = lb_tlb$p.value
	), digits = 4, caption = "TLB Ljung-Box test on residuals (lag = 10)"
)
```

```{r diag-tlb, echo=FALSE, fig.width=6, fig.height=5, results='hide'}
checkresiduals(tlb_model1)
```

# Forecasting Model & Assessment
```{r forecasting, echo=FALSE, fig.height=6, fig.width=6}
tfr_fc <- forecast(tfr_model1, h = 12)
tlb_fc <- forecast(tlb_model1, h = 12)

# TFR Forecasting
autoplot(tfr_fc) +
	autolayer(tfr_test, series = "Actual", colour = "red") +
	labs(
		title = "TFR Forecast (2013-2024) vs Actual",
		x = "Year", y = "TFR"
	) +
	theme_minimal()

# TLB Forecasting
autoplot(tlb_fc) +
	autolayer(tlb_test, series = "Actual", colour = "red") +
	labs(
		title = "TLB Forecast (2013-2024) vs Actual",
		x = "Year", y = "TLB"
	) +
	theme_minimal()

tfr_test_ts <- ts(tfr_test$TFR, start = 2013, frequency = 1)
tlb_test_ts <- ts(tlb_test$TLB, start = 2013, frequency = 1)

# Accuracy measurement
acc_tfr <- accuracy(tfr_fc, tfr_test_ts)
acc_tlb <- accuracy(tlb_fc, tlb_test_ts)

knitr::kable(acc_tfr, digits = 4, caption = "TFR forecast accuracy")
knitr::kable(acc_tlb, digits = 4, caption = "TLB forecast accuracy")
```

# Key Temporal Insights
The figures throughout this EDA show a sharp decline in both TLB and TFR post-1960s, which is indicative of Singapore's rapid demographic transformation. TLB drops from peaks around 1965 (~60,000 births) among the early baby booms, stabilising below 40,000 by the 2000s; TFR plummets from >4.5 in 1960 to a remaining fifth of this figure, with its notable troughs in the 1970s (during the presence of their two-limit policy) and 2020s (during the COVID-19 pandemic). Visual patterns indicate strong negative trends, potential cyclical behaviour, and increased volatility post-2010.

# Data Quality
The given Singapore Census CSV data is objectively clean; no NAs present, so no discrete approximations are made. However, a limitation is the lack of monthly granularity for our seasonality tests. This could be an inviting avenue to investigate, if present.

# Key Visual & Test Insights
Line plots are indicative of potential exponential decay (log-TLB linearises trend); differenced series highlight stationarity (expected ADF p<0.05 post-d=1). ACF shows high persistence (lag-1 ~ 0.9), weak cycles around 10-15 years (implementation of national policies, likely). In decomposition, trend is massive (~90% variance), with minimal seasonality in annuals (see Data Quality for relational limitations).

# Research Areas
1. What policy interventions over this sample period best explain the TFR/TLB deviations from smooth decline trend?
2. Do ARIMA/ETS models trained on the 1960-2012 slice accurately backcast 2013-2024 trends?
3. How do age-shifts and ethic gaps explain < 1.0 TFR?

# Report GitHub Repo
https://github.com/XCalibr3S/TSA_EDA.git 

