---
title: "Time Series Analysis - Singapore Total Fertility Rate\ 
  and Total Live Births"
author: "Hunter Holt"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: false
    toc_depth: 3
    number_sections: false
    df_print: kable
    highlight: tango
    latex_engine: pdflatex
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{ragged2e}
  - \usepackage{parskip}
  - \usepackage{setspace}
  - \usepackage{titlesec}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \setlength{\parindent}{0pt}
  - \setlength{\parskip}{6pt}
  - \titleformat{\section}{\normalfont\bfseries}{\thesection}{1em}{}
  - \titleformat{\subsection}{\normalfont\bfseries}{\thesubsection}{1em}{}
  - \titlespacing*{\section}{0pt}{12pt}{4pt}
  - \titlespacing*{\subsection}{0pt}{8pt}{2pt}
  - \renewcommand{\arraystretch}{1.15}
bibliography: references.bib
---

```{r setup, include=FALSE, warnings=FALSE}
knitr::opts_chunk$set(
	echo		= FALSE,
	warning	= FALSE,
	message		= FALSE,
	fig.align	= "center",
	fig.pos = "H",
  out.width = "85%"
)
```


```{r libs, include=FALSE}
library(tidyverse)	# for data wrangling & ggplot2
library(tseries)	# for adf & kpss tests
library(forecast)	# for acf/pacf, forecast & checkresiduals
library(tsibble)	# for tsibble data structure
library(knitr)
library(kableExtra)
```

---

# Introduction

This exploratory data analysis (EDA) investigates the temporal features of two significant population indicators for Singapore:

- **Total Fertility Rate (TFR):** the average number of children a woman would have over her lifetime at current age-specific fertility rates;
- **Total Live Births (TLB):** the total number of live births recognised per year.

The dates for the SingaporeCensus.csv reference dataset span from 1960 to 2024, sourced from the Singapore Department of Statistics. The objective is to recognise and analyse long-run trends, assess stationarity, and identify candidate time series models for a proposed training period 1960 - 2012 and a proposed forecasting performance assessed over the remaining 2013 - 2024 period.

# Data Loading & Reshaping

The given data is in a CSV file, which is optimal for computation. However, the data is in a 'wide format', where its temporal variables (i.e. years) are given as columns instead of years. We transpose the data and extract the two data columns we are interested in.

Below is the transposition framework of the data.

```{r clean-load-object, echo=FALSE}
raw <- read.csv("SingaporeCensus.csv", check.names = FALSE, na.strings = c("", "NA", "na"))

# Transpose
long <- raw |>
    rename(Series = DataSeries) |>
    pivot_longer(cols = -Series, names_to = "Year", values_to = "Value", 
                 names_transform = list(Year = as.integer)) |>
    filter(Year >= 1960, Year <= 2024)
	
# Extract TFR and TLB data
tfr_df <- long |>
	filter(trimws(Series) == "Total Fertility Rate (TFR)") |>
	arrange(Year)
tlb_df <- long |>
	filter(trimws(Series) == "Total Live-Births") |>
	arrange(Year)
	
# Verify 65 years covered per dataset
stopifnot(nrow(tfr_df) == 65, nrow(tlb_df) == 65)

# Create cleaned tsibbles
tfr_tsb <- tfr_df |>
	mutate(Year = as.integer(Year)) |>
	as_tsibble(index = Year) |>
	select(Year, TFR = Value)
	
tlb_tsb <- tlb_df |>
	mutate(Year = as.integer(Year)) |>
	as_tsibble(index = Year) |>
	select(Year, TLB = Value)

# Split tsibble for training & testing
tfr_train <- tfr_tsb |> filter(as.numeric(Year) <= 2012)
tlb_train <- tlb_tsb |> filter(as.numeric(Year) <= 2012)

tfr_test <- tfr_tsb |> filter(as.numeric(Year) >= 2013)
tlb_test <- tlb_tsb |> filter(as.numeric(Year) >= 2013)
```

```{r data-comp, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3.2, fig.width=8, out.width="100%"}
wide_sample <- data.frame(
	DataSeries = c("Total Fertility Rate (TFR)", "Total Live-Births (TLB)"),
	'1960' = c(5.76, 61775),
	'1961' = c(5.45, 62157),
	'...' = c("...", "..."),
	check.names = FALSE
)

long_sample <- data.frame(
	Series = c(
		"Total Fertility Rate (TFR)", "Total Fertility Rate (TFR)",
		"Total Live-Births", "Total Live-Births"
	),
	Year = c(1960, 1961, 1960, 1961),
	Value = c(5.76, 5.45, 61775, 62157)
)

wide_plot <- ggplot() +
	theme_void() + 
	annotate("text", x = 0, y = 0.9, label = "Wide format (raw CSV)", fontface = "bold", hjust = 0) +
	annotate("label", x = 0, y = 0.65, label = "DataSeries | 1960 | 1961 | ... | 2024\nTFR row\nTLB row",
	size = 3.2, label.size = 0.25, hjust = 0) +
	xlim(-0.1, 1) +
	ylim(0, 1.5)

arrow_plot <- ggplot() +
	theme_void() +
	annotate("text", x = 0.5, y = 0.65, label = "pivot_longer()\n+\nfilter()\n+\narrange()", fontface = "italic", color = "grey35") +
	xlim(0, 1) +
	ylim(0, 1.5)

long_plot <- ggplot() +
	theme_void() +
	annotate("text", x = 0, y = 0.95, label = "Long format (processed)", fontface = "bold", hjust = 0) +
	annotate("label", x = 0, y = 0.65, label = "Series | Year | Value\nTFR | 1960 | 5.76\nTFR | 1961 | 5.45\nTLB | 1960 | 61775\nTLB | 1961 | 62157",
	size = 3.2, label.size = 0.25, hjust = 0) +
	xlim(-0.1, 1) +
	ylim(0, 1.5)

(wide_plot | arrow_plot | long_plot) + plot_layout(widths = c(5, 5, 5))
```

```{r window_one, echo=FALSE}
tfr_ts <- ts(tfr_tsb$TFR, start = 1960, frequency = 1)
tlb_ts <- ts(tlb_tsb$TLB, start = 1960, frequency = 1)

tfr_train_ts <- window(tfr_ts, end = 2012)
tlb_train_ts <- window(tlb_ts, end = 2012)
```

# Data Quality & Summary
```{r summary, echo=FALSE}
summary_table <- tibble(
	Statistic = c("Min", "Q1", "Median", "Mean", "Q3", "Max", "NAs"),
	TFR = c(
		min(tfr_tsb$TFR, na.rm = TRUE), quantile(tfr_tsb$TFR, .25, na.rm = TRUE),
		median(tfr_tsb$TFR, na.rm = TRUE), mean(tfr_tsb$TFR, na.rm = TRUE),
		quantile(tfr_tsb$TFR, .75, na.rm = TRUE), max(tfr_tsb$TFR, na.rm = TRUE),
		sum(is.na(tfr_tsb$TFR))
	),
	TLB = c(
		min(tlb_tsb$TLB, na.rm = TRUE), quantile(tlb_tsb$TLB, .25, na.rm = TRUE),
		median(tlb_tsb$TLB, na.rm = TRUE), mean(tlb_tsb$TLB, na.rm = TRUE),
		quantile(tlb_tsb$TLB, .75, na.rm = TRUE), max(tlb_tsb$TLB, na.rm = TRUE),
		sum(is.na(tlb_tsb$TLB))
	)
)

knitr::kable(summary_table, digits = 2, caption = "Summary statistics for TFR and TLB (1960-2024)")
```

# Time Series Plots

Below, the dashed vertical line marks the train-test split at 2012.

```{r plot-tfr, echo=FALSE, fig.height=5, fig.width=4, out.width="60%"}
autoplot(tfr_tsb, TFR) +
	geom_vline(xintercept = 2012, linetype = "dashed", colour = "grey50") +
	annotate("text", x = 2013, y = max(tfr_tsb$TFR) * 0.95, label = "Test period ->", hjust = 0, size = 3.5, colour = "grey40") + labs(
		title = "Singapore Total Fertility Rate (1960-2024)",
		x = "Year",
		y = "TFR"
	) + theme_minimal()
```

```{r plot-tlb, echo=FALSE, fig.height=5, fig.width=4, out.width="60%"}
autoplot(tlb_tsb, TLB) +
	geom_vline(xintercept = 2012, linetype = "dashed", colour = "grey50") +
	annotate("text", x = 2013, y = max(tlb_tsb$TLB) * 0.95,
	label = "Test period ->", hjust = 0, size = 3.5, colour = "grey40") + labs(
		title = "Singapore Total Live Births (1960-2024)",
		x = "Year",
		y = "Live Births"
	) + theme_minimal()
```

# Temporal Feature Evaluation

## Log Transformation Assessment

```{r log-transform, echo=FALSE, fig.width=6, fig.height=4.5}
par(mfrow = c(2, 2), mar = c(4, 4, 2.5, 1))
plot(tfr_ts, main = "TFR - Raw", ylab = "TFR")
plot(log(tfr_ts), main = "TFR - Log", ylab = "log(TFR)")
plot(tlb_ts, main = "TLB - Raw", ylab = "Live Births")
plot(log(tlb_ts), main = "TLB - Log", ylab = "log(TLB)")

par(mfrow = c(1, 1))
```

\newpage
# Stationarity Testing

## Raw Series

```{r stationarity-raw, echo=FALSE}
adf_tfr <- adf.test(tfr_train_ts)
kpss_tfr <- kpss.test(tfr_train_ts)

adf_tlb <- adf.test(tlb_train_ts)
kpss_tlb <- kpss.test(tlb_train_ts)

stationarity_raw <- tibble(
	Series = c("TFR", "TLB"),
	`ADF p-value` = c(adf_tfr$p.value, adf_tlb$p.value),
	`KPSS p-value` = c(kpss_tfr$p.value, kpss_tlb$p.value)
)

knitr::kable(stationarity_raw, digits = 4,
	caption = "Stationarity tests - raw training series (1960-2012)")
```

## First Difference Series

```{r stationarity-diff, echo=FALSE}
adf_tfr_d <- adf.test(diff(tfr_train_ts))
kpss_tfr_d <- kpss.test(diff(tfr_train_ts))

adf_tlb_d <- adf.test(diff(tlb_train_ts))
kpss_tlb_d <- kpss.test(diff(tlb_train_ts))

stationarity_diff <- tibble(
	Series = c("$\\Delta$TFR", "$\\Delta$TLB"),
	`ADF p-value` = c(adf_tfr_d$p.value, adf_tlb_d$p.value),
	`KPSS p-value` = c(kpss_tfr_d$p.value, kpss_tlb_d$p.value)
)

knitr::kable(stationarity_diff, digits = 4,
	escape = FALSE,
	caption = "Stationarity tests - first-differenced training series")
```

\newpage

# ACF & PACF Analysis

```{r acf-pacf-raw, echo=FALSE, fig.width = 6, fig.height = 4.5}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

Acf(tfr_train_ts, main = "TFR - ACF")
Pacf(tfr_train_ts, main = "TFR - PACF")
Acf(tlb_train_ts, main = "TLB - ACF")
Pacf(tlb_train_ts, main = "TLB - PACF")

par(mfrow = c(1, 1))
```

```{r acf-pacf-diff, echo=FALSE, fig.width = 6, fig.height = 4.5}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

Acf(diff(tfr_train_ts), main = expression(Delta * TFR ~ "- ACF"))
Pacf(diff(tfr_train_ts), main = expression(Delta * TFR ~ "- PACF"))
Acf(diff(tlb_train_ts), main = expression(Delta * TLB ~ "- ACF"))
Pacf(diff(tlb_train_ts), main = expression(Delta * TLB ~ "- PACF"))

par(mfrow = c(1, 1))
```

# Decomposition
```{r decompose-tfr, echo=FALSE, fig.width=7, fig.height=5}
tfr_trend <- stats::filter(tfr_train_ts, filter = rep(1/5, 5), sides = 2)

plot(tfr_train_ts, main = "TFR - Trend (5-year MA)", ylab = "TFR", col = "grey60")
lines(tfr_trend, col = "steelblue", lwd = 2)
```

```{r decompose-tlb, echo=FALSE, fig.width=7, fig.height=5}
tlb_trend <- stats::filter(tlb_train_ts, filter = rep(1/5, 5), sides = 2)

plot(tlb_train_ts, main = "TLB - Trend (5-year MA)", ylab = "TLB", col = "grey60")
lines(tlb_trend, col = "steelblue", lwd = 2)
```

# Candidate Model Fitting

## TFR - Candidate Model

```{r model-tfr, echo=FALSE, fig.width=6, fig.height=5, results='hide'}
tfr_model1 <- Arima(tfr_train_ts, order = c(0, 0, 0)) # TBD
```

```{r model-tfr-results, echo=FALSE}
lb_tfr <- Box.test(residuals(tfr_model1), lag = 10, type = "Ljung-Box", fitdf = 0)

knitr::kable(
	accuracy(tfr_model1) |> as.data.frame(),
	digits = 4, caption = "TFR training set error measures"
)

knitr::kable(
	tibble(
		'Q*' = lb_tfr$statistic,
		df = lb_tfr$parameter,
		'p-value' = lb_tfr$p.value
	), digits = 4, caption = "TFR Ljung-Box test on residuals (lag = 10)"
)
```

```{r diag-tfr, echo=FALSE, fig.width=6, fig.height=5, results='hide'}
checkresiduals(tfr_model1)
```

\newpage
## TLB - Candidate Model

```{r model-tlb, echo=FALSE, fig.width=6, fig.height=5, results='hide'}
tlb_model1 <- Arima(tlb_train_ts, order = c(0, 0, 0)) # TBD
```

```{r model-tlb-results, echo=FALSE}
lb_tlb <- Box.test(residuals(tlb_model1), lag = 10, type = "Ljung-Box", fitdf = 0)

knitr::kable(
	accuracy(tlb_model1) |> as.data.frame(),
	digits = 4, caption = "TLB training set error measures"
)

knitr::kable(
	tibble(
		'Q*' = lb_tlb$statistic,
		df = lb_tlb$parameter,
		'p-value' = lb_tlb$p.value
	), digits = 4, caption = "TLB Ljung-Box test on residuals (lag = 10)"
)
```

```{r diag-tlb, echo=FALSE, fig.width=6, fig.height=5, results='hide'}
checkresiduals(tlb_model1)
```

# Forecasting Model & Assessment
```{r forecasting, echo=FALSE, fig.height=6, fig.width=6}
tfr_fc <- forecast(tfr_model1, h = 12)
tlb_fc <- forecast(tlb_model1, h = 12)

# TFR Forecasting
autoplot(tfr_fc) +
	autolayer(tfr_test, series = "Actual", colour = "red") +
	labs(
		title = "TFR Forecast (2013-2024) vs Actual",
		x = "Year", y = "TFR"
	) +
	theme_minimal()

# TLB Forecasting
autoplot(tlb_fc) +
	autolayer(tlb_test, series = "Actual", colour = "red") +
	labs(
		title = "TLB Forecast (2013-2024) vs Actual",
		x = "Year", y = "TLB"
	) +
	theme_minimal()

tfr_test_ts <- ts(tfr_test$TFR, start = 2013, frequency = 1)
tlb_test_ts <- ts(tlb_test$TLB, start = 2013, frequency = 1)

# Accuracy measurement
acc_tfr <- accuracy(tfr_fc, tfr_test_ts)
acc_tlb <- accuracy(tlb_fc, tlb_test_ts)

knitr::kable(acc_tfr, digits = 4, caption = "TFR forecast accuracy")
knitr::kable(acc_tlb, digits = 4, caption = "TLB forecast accuracy")
```

# Key Temporal Insights
The figures throughout this EDA show a sharp decline in both TLB and TFR post-1960s, which is indicative of Singapore's rapid demographic transformation. TLB drops from peaks around 1965 (~60,000 births) among the early baby booms, stabilising below 40,000 by the 2000s; TFR plummets from >4.5 in 1960 to a remaining fifth of this figure, with its notable troughs in the 1970s (during the presence of their two-limit policy) and 2020s (during the COVID-19 pandemic). Visual patterns indicate strong negative trends, potential cyclical behaviour, and increased volatility post-2010.

# Data Quality
The given Singapore Census CSV data is objectively clean; no NAs present, so no discrete approximations are made. However, a limitation is the lack of monthly granularity for our seasonality tests. This could be an inviting avenue to investigate, if present.

# Key Visual & Test Insights
Line plots are indicative of potential exponential decay (log-TLB linearises trend); differenced series highlight stationarity (expected ADF p<0.05 post-d=1). ACF shows high persistence (lag-1 ~ 0.9), weak cycles around 10-15 years (implementation of national policies, likely). In decomposition, trend is massive (~90% variance), with minimal seasonality in annuals (see Data Quality for relational limitations).

# Research Areas
1. What policy interventions over this sample period best explain the TFR/TLB deviations from smooth decline trend?
2. Do ARIMA/ETS models trained on the 1960-2012 slice accurately backcast 2013-2024 trends?
3. How do age-shifts and ethic gaps explain < 1.0 TFR?

# Report GitHub Repo
https://github.com/XCalibr3S/TSA_EDA.git 

