---
title: "Time Series Analysis - Singapore Total Fertility Rate\ 
  and Total Live Births"
author: "Hunter Holt"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: false
    toc_depth: 3
    number_sections: false
    df_print: kable
    highlight: tango
    latex_engine: pdflatex
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{ragged2e}
  - \usepackage{parskip}
  - \usepackage{setspace}
  - \usepackage{titlesec}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \setlength{\parindent}{0pt}
  - \setlength{\parskip}{6pt}
  - \titleformat{\section}{\normalfont\bfseries}{\thesection}{1em}{}
  - \titleformat{\subsection}{\normalfont\bfseries}{\thesubsection}{1em}{}
  - \titlespacing*{\section}{0pt}{12pt}{4pt}
  - \titlespacing*{\subsection}{0pt}{8pt}{2pt}
  - \renewcommand{\arraystretch}{1.15}
bibliography: references.bib
---

```{r setup, include=FALSE, warnings=FALSE}
knitr::opts_chunk$set(
	echo		= FALSE,
	warning	= FALSE,
	message		= FALSE,
	fig.align	= "center",
	fig.pos = "H",
  out.width = "85%"
)
```


```{r libs, include=FALSE}
library(tidyverse)	# for data wrangling & ggplot2
library(tseries)	# for adf & kpss tests
library(forecast)	# for acf/pacf, forecast & checkresiduals
library(tsibble)	# for tsibble data structure
library(knitr)
library(kableExtra)
```

---

\begin{center}
\large\textbf{Abstract}
\end{center}

Over the course of the last century, Singapore's demographic transition remains as one of the most rapid ever recorded, with the Total Fertility Rate (TFR) declining from 5.76 in 1960 to less than 1.0 by 2024. This paper features a complete time series analysis of two interrelated population metrics: TFR and Total Live Births (TLB). These are scraped from the Singapore Department of Statistics' annual data, ranging from 1968 to 2024. After an established Data Analysis Lifecycle, paired with the Box-Jenkins identification cycle, a log-differencing transformation is required to achieve stationarity. This is also furthermore confirmed by ADF and KPSS tests. Autocorrelation (ACF) and partial autocorrelation (ACF) functions applied to our stationary series prompts for an ARIMA$(p, 1, q)$ candidate model(s) fitted with a logarithmic Box-Cox transform. Model selection verification using AICc identifies the specification with best fit for each series; residual diagnostics aim to replicate white noise behaviour. Out-of-sample forecasts over a subsample of 2013--2024 are evaluated against observed values, with performance assessed via RMSE, MAE, MAPE, and Theil's U. The results post-evaluation confirm the particular difficulty of accurately forecasting sub-replacement fertility prediction mapping, likely due to structural policy interventions within Singapore and/or the impact of COVID-19. This motivates additional extension processes including ARIMA-GARCH and state-space modelling.

*Keywords:* Total Fertility Rate; Total Live Births; ARIMA; Box-Jenkins; Stationarity; Singapore

\vspace{10pt}
\hrule
\vspace{8pt}

---

# Introduction

- Demographic transition
- Discuss variables
- Discuss significant change in variables, and why. EDA
- Introduce research questions: 'which policy interventions over 1960--2024 justify the significant structural deviations in TFR and TLB?', 'can ARIMA models trained on the 1960--2012 window project the 2013--2024 period?', 'what (and how do) demographic mechanisms account for the recent significant drop in TFR?'
- State IED Box-Jenkins cycle as diagnostic tool

# Data Loading & Reshaping

```{r load-data}
raw <- read.csv("SingaporeCensus.csv", 
                check.names = FALSE, 
                na.strings = c("", 
                               "NA", 
                               "na")
)

# Transpose
long <- raw |>
    rename(Series = DataSeries) |>
    pivot_longer(cols = -Series,
                 names_to = "Year",
                 values_to = "Value", 
                 names_transform = list(Year = as.integer)
                 ) |>
    filter(Year >= 1960, Year <= 2024)
	
# Extract TFR and TLB data
tfr_df <- long |>
	filter(trimws(Series) == "Total Fertility Rate (TFR)") |>
	arrange(Year)
tlb_df <- long |>
	filter(trimws(Series) == "Total Live-Births") |>
	arrange(Year)
	
# Verify 65 years covered per dataset
stopifnot(nrow(tfr_df) == 65, nrow(tlb_df) == 65)

# Create cleaned tsibbles
tfr_tsb <- ts(tfr_df$Value, start = 1960, frequency = 1)
tlb_tsb <- ts(tlb_df$Value, start = 1960, frequency = 1)

# Split tsibble for training & testing
tfr_train <- window(tfr_tsb, end = 2012)
tfr_test <- window(tfr_tsb, start = 2013)
tlb_train <- window(tlb_tsb, end = 2012)
tlb_train <- window(tlb_tsb, start = 2013)
```

The data used throughout this report are sourced from *SingaporeCensus.csv*, a population census published by the Singapore Department of Statistics.
The raw comma-separated-value file is stored in wide format, i.e. the years 1960--2024 are column headers. Two rows are extracted -- one for TFR and one for TLB -- and reshaped to long format by using `pivot_longer()`. The resulting cleaned dataset contains 65 annual observations per series with no missing values.

For forecasting evaluation, the full series is partitioned into a training windows of $n = 53$ observations (from 1960 to 2012) and a complementary windows of $h = 12$ observations (2013 to 2024).

```{r data-comp, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3.2, fig.width=8, out.width="100%"}
wide_sample <- data.frame(
	DataSeries = c("Total Fertility Rate (TFR)", "Total Live-Births (TLB)"),
	'1960' = c(5.76, 61775),
	'1961' = c(5.45, 62157),
	'...' = c("...", "..."),
	check.names = FALSE
)

long_sample <- data.frame(
	Series = c(
		"Total Fertility Rate (TFR)", "Total Fertility Rate (TFR)",
		"Total Live-Births", "Total Live-Births"
	),
	Year = c(1960, 1961, 1960, 1961),
	Value = c(5.76, 5.45, 61775, 62157)
)

wide_plot <- ggplot() +
	theme_void() + 
	annotate("text", x = 0, y = 0.9, label = "Wide format (raw CSV)", fontface = "bold", hjust = 0) +
	annotate("label", x = 0, y = 0.65, label = "DataSeries | 1960 | 1961 | ... | 2024\nTFR row\nTLB row",
	size = 3.2, label.size = 0.25, hjust = 0) +
	xlim(-0.1, 1) +
	ylim(0, 1.5)

arrow_plot <- ggplot() +
	theme_void() +
	annotate("text", x = 0.5, y = 0.65, label = "pivot_longer()\n+\nfilter()\n+\narrange()", fontface = "italic", color = "grey35") +
	xlim(0, 1) +
	ylim(0, 1.5)

long_plot <- ggplot() +
	theme_void() +
	annotate("text", x = 0, y = 0.95, label = "Long format (processed)", fontface = "bold", hjust = 0) +
	annotate("label", x = 0, y = 0.65, label = "Series | Year | Value\nTFR | 1960 | 5.76\nTFR | 1961 | 5.45\nTLB | 1960 | 61775\nTLB | 1961 | 62157",
	size = 3.2, label.size = 0.25, hjust = 0) +
	xlim(-0.1, 1) +
	ylim(0, 1.5)

(wide_plot | arrow_plot | long_plot) + plot_layout(widths = c(5, 5, 5))
```

```{r window_one, echo=FALSE}
tfr_ts <- ts(tfr_tsb$TFR, start = 1960, frequency = 1)
tlb_ts <- ts(tlb_tsb$TLB, start = 1960, frequency = 1)

tfr_train_ts <- window(tfr_ts, end = 2012)
tlb_train_ts <- window(tlb_ts, end = 2012)
```

# Data Quality & Summary
```{r summary, echo=FALSE}
summary_table <- tibble(
	Statistic = c("Min", "Q1", "Median", "Mean", "Q3", "Max", "NAs"),
	TFR = c(
		min(tfr_tsb$TFR, na.rm = TRUE), quantile(tfr_tsb$TFR, .25, na.rm = TRUE),
		median(tfr_tsb$TFR, na.rm = TRUE), mean(tfr_tsb$TFR, na.rm = TRUE),
		quantile(tfr_tsb$TFR, .75, na.rm = TRUE), max(tfr_tsb$TFR, na.rm = TRUE),
		sum(is.na(tfr_tsb$TFR))
	),
	TLB = c(
		min(tlb_tsb$TLB, na.rm = TRUE), quantile(tlb_tsb$TLB, .25, na.rm = TRUE),
		median(tlb_tsb$TLB, na.rm = TRUE), mean(tlb_tsb$TLB, na.rm = TRUE),
		quantile(tlb_tsb$TLB, .75, na.rm = TRUE), max(tlb_tsb$TLB, na.rm = TRUE),
		sum(is.na(tlb_tsb$TLB))
	)
)

knitr::kable(summary_table, digits = 2, caption = "Summary statistics for TFR and TLB (1960-2024)")
```

# Time Series Plots

Below, the dashed vertical line marks the train-test split at 2012.

```{r plot-tfr, echo=FALSE, fig.height=5, fig.width=4, out.width="60%"}
autoplot(tfr_tsb, TFR) +
	geom_vline(xintercept = 2012, linetype = "dashed", colour = "grey50") +
	annotate("text", x = 2013, y = max(tfr_tsb$TFR) * 0.95, label = "Test period ->", hjust = 0, size = 3.5, colour = "grey40") + labs(
		title = "Singapore Total Fertility Rate (1960-2024)",
		x = "Year",
		y = "TFR"
	) + theme_minimal()
```

```{r plot-tlb, echo=FALSE, fig.height=5, fig.width=4, out.width="60%"}
autoplot(tlb_tsb, TLB) +
	geom_vline(xintercept = 2012, linetype = "dashed", colour = "grey50") +
	annotate("text", x = 2013, y = max(tlb_tsb$TLB) * 0.95,
	label = "Test period ->", hjust = 0, size = 3.5, colour = "grey40") + labs(
		title = "Singapore Total Live Births (1960-2024)",
		x = "Year",
		y = "Live Births"
	) + theme_minimal()
```

# Temporal Feature Evaluation

## Log Transformation Assessment

```{r log-transform, echo=FALSE, fig.width=6, fig.height=4.5}
par(mfrow = c(2, 2), mar = c(4, 4, 2.5, 1))
plot(tfr_ts, main = "TFR - Raw", ylab = "TFR")
plot(log(tfr_ts), main = "TFR - Log", ylab = "log(TFR)")
plot(tlb_ts, main = "TLB - Raw", ylab = "Live Births")
plot(log(tlb_ts), main = "TLB - Log", ylab = "log(TLB)")

par(mfrow = c(1, 1))
```

\newpage
# Stationarity Testing

## Raw Series

```{r stationarity-raw, echo=FALSE}
adf_tfr <- adf.test(tfr_train_ts)
kpss_tfr <- kpss.test(tfr_train_ts)

adf_tlb <- adf.test(tlb_train_ts)
kpss_tlb <- kpss.test(tlb_train_ts)

stationarity_raw <- tibble(
	Series = c("TFR", "TLB"),
	`ADF p-value` = c(adf_tfr$p.value, adf_tlb$p.value),
	`KPSS p-value` = c(kpss_tfr$p.value, kpss_tlb$p.value)
)

knitr::kable(stationarity_raw, digits = 4,
	caption = "Stationarity tests - raw training series (1960-2012)")
```

## First Difference Series

```{r stationarity-diff, echo=FALSE}
adf_tfr_d <- adf.test(diff(tfr_train_ts))
kpss_tfr_d <- kpss.test(diff(tfr_train_ts))

adf_tlb_d <- adf.test(diff(tlb_train_ts))
kpss_tlb_d <- kpss.test(diff(tlb_train_ts))

stationarity_diff <- tibble(
	Series = c("$\\Delta$TFR", "$\\Delta$TLB"),
	`ADF p-value` = c(adf_tfr_d$p.value, adf_tlb_d$p.value),
	`KPSS p-value` = c(kpss_tfr_d$p.value, kpss_tlb_d$p.value)
)

knitr::kable(stationarity_diff, digits = 4,
	escape = FALSE,
	caption = "Stationarity tests - first-differenced training series")
```

\newpage

# ACF & PACF Analysis

```{r acf-pacf-raw, echo=FALSE, fig.width = 6, fig.height = 4.5}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

Acf(tfr_train_ts, main = "TFR - ACF")
Pacf(tfr_train_ts, main = "TFR - PACF")
Acf(tlb_train_ts, main = "TLB - ACF")
Pacf(tlb_train_ts, main = "TLB - PACF")

par(mfrow = c(1, 1))
```

```{r acf-pacf-diff, echo=FALSE, fig.width = 6, fig.height = 4.5}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

Acf(diff(tfr_train_ts), main = expression(Delta * TFR ~ "- ACF"))
Pacf(diff(tfr_train_ts), main = expression(Delta * TFR ~ "- PACF"))
Acf(diff(tlb_train_ts), main = expression(Delta * TLB ~ "- ACF"))
Pacf(diff(tlb_train_ts), main = expression(Delta * TLB ~ "- PACF"))

par(mfrow = c(1, 1))
```

# Decomposition
```{r decompose-tfr, echo=FALSE, fig.width=7, fig.height=5}
tfr_trend <- stats::filter(tfr_train_ts, filter = rep(1/5, 5), sides = 2)

plot(tfr_train_ts, main = "TFR - Trend (5-year MA)", ylab = "TFR", col = "grey60")
lines(tfr_trend, col = "steelblue", lwd = 2)
```

```{r decompose-tlb, echo=FALSE, fig.width=7, fig.height=5}
tlb_trend <- stats::filter(tlb_train_ts, filter = rep(1/5, 5), sides = 2)

plot(tlb_train_ts, main = "TLB - Trend (5-year MA)", ylab = "TLB", col = "grey60")
lines(tlb_trend, col = "steelblue", lwd = 2)
```

# Candidate Model Fitting

## TFR - Candidate Model

```{r model-tfr, echo=FALSE, fig.width=6, fig.height=5, results='hide'}
tfr_model1 <- Arima(tfr_train_ts, order = c(0, 0, 0)) # TBD
```

```{r model-tfr-results, echo=FALSE}
lb_tfr <- Box.test(residuals(tfr_model1), lag = 10, type = "Ljung-Box", fitdf = 0)

knitr::kable(
	accuracy(tfr_model1) |> as.data.frame(),
	digits = 4, caption = "TFR training set error measures"
)

knitr::kable(
	tibble(
		'Q*' = lb_tfr$statistic,
		df = lb_tfr$parameter,
		'p-value' = lb_tfr$p.value
	), digits = 4, caption = "TFR Ljung-Box test on residuals (lag = 10)"
)
```

```{r diag-tfr, echo=FALSE, fig.width=6, fig.height=5, results='hide'}
checkresiduals(tfr_model1)
```

\newpage
## TLB - Candidate Model

```{r model-tlb, echo=FALSE, fig.width=6, fig.height=5, results='hide'}
tlb_model1 <- Arima(tlb_train_ts, order = c(0, 0, 0)) # TBD
```

```{r model-tlb-results, echo=FALSE}
lb_tlb <- Box.test(residuals(tlb_model1), lag = 10, type = "Ljung-Box", fitdf = 0)

knitr::kable(
	accuracy(tlb_model1) |> as.data.frame(),
	digits = 4, caption = "TLB training set error measures"
)

knitr::kable(
	tibble(
		'Q*' = lb_tlb$statistic,
		df = lb_tlb$parameter,
		'p-value' = lb_tlb$p.value
	), digits = 4, caption = "TLB Ljung-Box test on residuals (lag = 10)"
)
```

```{r diag-tlb, echo=FALSE, fig.width=6, fig.height=5, results='hide'}
checkresiduals(tlb_model1)
```

# Forecasting Model & Assessment
```{r forecasting, echo=FALSE, fig.height=6, fig.width=6}
tfr_fc <- forecast(tfr_model1, h = 12)
tlb_fc <- forecast(tlb_model1, h = 12)

# TFR Forecasting
autoplot(tfr_fc) +
	autolayer(tfr_test, series = "Actual", colour = "red") +
	labs(
		title = "TFR Forecast (2013-2024) vs Actual",
		x = "Year", y = "TFR"
	) +
	theme_minimal()

# TLB Forecasting
autoplot(tlb_fc) +
	autolayer(tlb_test, series = "Actual", colour = "red") +
	labs(
		title = "TLB Forecast (2013-2024) vs Actual",
		x = "Year", y = "TLB"
	) +
	theme_minimal()

tfr_test_ts <- ts(tfr_test$TFR, start = 2013, frequency = 1)
tlb_test_ts <- ts(tlb_test$TLB, start = 2013, frequency = 1)

# Accuracy measurement
acc_tfr <- accuracy(tfr_fc, tfr_test_ts)
acc_tlb <- accuracy(tlb_fc, tlb_test_ts)

knitr::kable(acc_tfr, digits = 4, caption = "TFR forecast accuracy")
knitr::kable(acc_tlb, digits = 4, caption = "TLB forecast accuracy")
```

# Key Temporal Insights
The figures throughout this EDA show a sharp decline in both TLB and TFR post-1960s, which is indicative of Singapore's rapid demographic transformation. TLB drops from peaks around 1965 (~60,000 births) among the early baby booms, stabilising below 40,000 by the 2000s; TFR plummets from >4.5 in 1960 to a remaining fifth of this figure, with its notable troughs in the 1970s (during the presence of their two-limit policy) and 2020s (during the COVID-19 pandemic). Visual patterns indicate strong negative trends, potential cyclical behaviour, and increased volatility post-2010.

# Data Quality
The given Singapore Census CSV data is objectively clean; no NAs present, so no discrete approximations are made. However, a limitation is the lack of monthly granularity for our seasonality tests. This could be an inviting avenue to investigate, if present.

# Key Visual & Test Insights
Line plots are indicative of potential exponential decay (log-TLB linearises trend); differenced series highlight stationarity (expected ADF p<0.05 post-d=1). ACF shows high persistence (lag-1 ~ 0.9), weak cycles around 10-15 years (implementation of national policies, likely). In decomposition, trend is massive (~90% variance), with minimal seasonality in annuals (see Data Quality for relational limitations).

# Research Areas
1. What policy interventions over this sample period best explain the TFR/TLB deviations from smooth decline trend?
2. Do ARIMA/ETS models trained on the 1960-2012 slice accurately backcast 2013-2024 trends?
3. How do age-shifts and ethic gaps explain < 1.0 TFR?

# Report GitHub Repo
https://github.com/XCalibr3S/TSA_EDA.git 

